Density-dependent selection and the limits of relative fitness

### 2018-09-01 manuscript.tex
l

The properties of density-dependent selection in our model are strikingly different from the classical literature discussed above. The strong connection between crowding and selection for greater equilibrium density is broken: selection need not affect density at all. And when it does, the density-independent discrete-time selection equation \eqref{eq:canonicaldisc} is almost exact even for strong selection, provided that any changes in density are driven only by selection (as opposed to large deviations from demogaphic equilibrium), and that selection occurs on only one of the traits $b$, $c$, or $d$. On the flip side, the constant relative fitness approximation fails when strong selection acts concurrently on two or more of these traits, or when the population is far from demographic equilibrium.

@@

We assume that adults cannot be ousted by juveniles, so that recruitment to adulthood occurs exclusively in unoccupied territories. When multiple propagules land on the same unoccupied territory, the winner is determined by lottery competition: type $i$ wins a territory with probability $c_i x_i/\sum_i c_i x_i$, where $c_i$ is a constant representing relative competitive ability (Fig. \ref{fig:lottery}). Since the expected fraction of unoccupied territories with propagule composition $x_1,\ldots,x_G$ is $p_1(x_1)\cdots p_G(x_G)$ where $G$ is the number of types present, and type $i$ is expected to win a proportion $c_i x_i/\sum_i c_i x_i$ of these, type $i$'s expected territorial acquisition is given by
\begin{equation}
\Delta_+ n_i=U\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_i c_i x_i} p_1(x_1)\cdots p_G(x_G). \label{eq:growthsumuncoupled}
\end{equation}
Here the sum only includes territories with at least one propagule present. Note that $\Delta_+ n_i$ denotes the \textit{expected} territorial acquisition. Fluctuations about $\Delta_+ n_i$ (i.e.~drift) will not be analyzed in this manuscript. Note that drift can become important if $U$ is not sufficiently large even though $n_i$ and $T$ are large (by assumption); we do not consider this scenario on biological grounds, since it implies negligible population turnover.

Adult mortality occurs after lottery recruitment at a constant, type-specific per-capita rate $d_i\geq 1$, and can affect adults recruited in the current iteration, such that the new abundance at the end of the iteration is $(n_i+\Delta_+ n_i)/d_i$ (Fig.~\ref{fig:lottery}). In terms of absolute fitness, this can be written as
\begin{equation}
W_i=\frac{1}{d_i}\left(1+\frac{\Delta_+ n_i}{n_i}\right). \label{eq:absfit}
\end{equation}
Here $\frac{\Delta_+ n_i}{n_i}$ is the per-capita rate of territorial acquisition, and $1/d_i$ is the fraction of type $i$ adults surviving to the next iteration.

@@

Selection in our variable-density lottery model is in general density-dependent, by which we mean that the discrete-time selection factor $(W_2-W_1)/\overline{W}$ from Eq.~\eqref{eq:canonicaldisc} may depend on $N$. More specifically, as we show below, $b$- and $c$- selection are density-dependent, but $d$-selection is not. Note that density-dependent selection is sometimes taken to mean a qualitative change in which types are fitter than others at different densities \citep{travis_2013}. While reversal in the order of fitnesses and co-existence driven by density-regulation are possible in our variable-density lottery (a special case of the competition-colonization trade-off; \citealt{levins_71,tilman_94,bolker_99}), questions related to co-existence are tangential to our aims and will not be pursued further here.

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{strengthofselection.pdf}
\caption{\label{fig:strengthofselection} Change in the selection coefficient between the beginning and end of a sweep of a type that experiences proportionally $1-\epsilon$ fold fewer crowding-induced deaths. The population is in demographic equilibrium at the start and end of the sweep.}
\end{figure}

In our variable density lottery, $b$ regulates density and is subject to density-dependent selection, yet $b$-sweeps are qualitatively different from $\delta$ sweeps in the above example. Greater $b$ means not only that more propagules contest the available territories, but also that a greater fraction of unoccupied territories receive propagules. Together, the net density-dependent effect on $b$-selection is negligible: in a single-type equilibrium we have $W_i=1$ and $b_i/\overline{b}=1$, and hence the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ in Eq.~\eqref{eq:masterequalc} has the same value $d_i-1$ at the beginning and end of a $b$-sweep (recall that $\frac{W_2-W_1}{\overline{W}}=\frac{f(\overline{b},N)}{1+f(\overline{b},N)}\frac{b_2-b_1}{\overline{b}}$ for $b$-selection). During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix C for details). Since selection must already be strong for a $\delta$-sweep to invalidate Eq.~\eqref{eq:canonicalcont}, the density-independent model applies almost exactly for equilibrium $b$-sweeps (Fig.~\ref{fig:bsweep}).

@@

In practice, traits like $b$, $c$ and $d$ will often have pleiotropic interactions which mean that adaptive sweeps will behave similarly to the familiar ``$\delta$ sweeps'' (``Density regulating traits under strong selection''). Thus, our analysis of the variable density lottery does not suggest that Eq.~\eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc} should apply more broadly than previously thought. Rather, $b$, $c$ and $d$ represent a possible set of idealized fitness components mediating the interplay between selection and density in density-regulated populations. The conceptual distinctions they highlight will hopefully be useful regardless of the biological realism of the variable density lottery model.

@@

In the analysis presented here we have restricted our attention to selection in demographically stable populations. The largest deviations from the approximation of density-independent selection (as represented by Eqs.~\eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc}) will likely occur in populations far from demographic equilibrium e.g.~as a result of a temporally-variable environment. This is because extremely strong selection to change population density by an amount comparable to environmental varability (see Fig.~\ref{fig:strengthofselection}). By contrast, temporally-variable environments can dramatically alter frequency trajectories for individual sweeps (e.g. Fig.~9.5 in \cite{otto_2011}; Fig.~5 in \cite{mallet_2012}), as well as the long-term outcomes of selection \citep{lande_2009}.

@@

This pathological behavior occurs because the expected total density of propagules in the respective groups of territories are different; $\langle X \rangle_{p({\mathbf x}|x_1=1,X_1\geq 1)}\approx l_2+1 > \langle X \rangle_{p({\mathbf x}|x_2\geq 2)}=\approx l_2 $. As a result, the rare type's behavior is approximated as though it experiences more intense lottery competition than the common type, which cannot be the case since the two types are identical. The effective means must thus be taken in a way that ensures that the expected total propagule density is the same in Eq.~\eqref{eq:meanfieldr} and Eq.~\eqref{eq:meanfielda}.

@@

Having derived the approximation Eq.~\eqref{eq:master}, we now evaluate its domain of validity. Eq.~\eqref{eq:master} relies on ignoring the fluctuations in $x_i$ and $x_j$, such that we can replace them with constant effective mean values. To justify this, we show that the standard deviations $\sigma_{p({\mathbf x}|x_i=1,X_i\geq 1)}(\sum_{j\neq i} c_j x_j)$ and $\sigma_{p({\mathbf x}|x_i\geq 2)}(\sum_j c_j x_j)$ are small compared to the corresponding means $\langle\sum_{j\neq i} c_j x_j\rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ and $\langle\sum_j c_j x_j\rangle_{p({\mathbf x}|x_i\geq 2)}$ in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda}. This result means that using the exact distributions $p({\mathbf x}|x_i=1,X_i\geq 1)$ and $p({\mathbf x}|x_i\geq 2)$ for the effective means would produce an accurate approximation of the components in \eqref{eq:delt_decomp} (though, as we have seen, not a consistent one). It is then clear that the effective means derived in the previous section will also give an accurate approximation since their magnitudes are similar to the exact means; this is obvious from the fact that the expected total number of propagules is of order $\min\{L,2\}$ in both cases.

>>>>

The properties of density-dependent selection in our model are strikingly different from the classical literature discussed above. The strong connection between crowding and selection for greater equilibrium density is broken: selection need not affect density at all. And when it does, the density-independent discrete-time selection equation \eqref{eq:canonicaldisc} is almost exact even for strong selection, provided that any changes in density are driven only by selection (as opposed to large deviations from demographic equilibrium), and that selection occurs on only one of the traits $b$, $c$, or $d$. On the flip side, the constant relative fitness approximation fails when strong selection acts concurrently on two or more of these traits, or when the population is far from demographic equilibrium.

@@

We assume that adults cannot be ousted by juveniles, so that recruitment to adulthood occurs exclusively in unoccupied territories. When multiple propagules land on the same unoccupied territory, the winner is determined by lottery competition: type $i$ wins a territory with probability $c_i x_i/\sum_i c_i x_i$, where $c_i$ is a constant representing relative competitive ability (Fig. \ref{fig:lottery}). Since the expected fraction of unoccupied territories with propagule composition $x_1,\ldots,x_G$ is $p_1(x_1)\cdots p_G(x_G)$ where $G$ is the number of types present, and type $i$ is expected to win a proportion $c_i x_i/\sum_i c_i x_i$ of these, type $i$'s expected territorial acquisition is given by
\begin{equation}
\Delta_+ n_i=U\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_i c_i x_i} p_1(x_1)\cdots p_G(x_G). \label{eq:growthsumuncoupled}
\end{equation}
Here the sum only includes territories with at least one propagule present. Note that $\Delta_+ n_i$ denotes the expected territorial acquisition. Fluctuations about $\Delta_+ n_i$ (i.e.~drift) will not be analyzed in this manuscript. Note that drift can become important if $U$ is not sufficiently large even though $n_i$ and $T$ are large (by assumption); we do not consider this scenario on biological grounds, since it implies negligible population turnover.

Adult mortality occurs after lottery recruitment at a constant, type-specific per-capita rate $d_i\geq 1$, and can affect adults recruited in the current iteration, such that the new abundance at the end of the iteration is $(n_i+\Delta_+ n_i)/d_i$ (Fig.~\ref{fig:lottery}). In terms of absolute fitness, this can be written as
\begin{equation}
W_i=\frac{1}{d_i}\left(1+\frac{\Delta_+ n_i}{n_i}\right). \label{eq:absfit}
\end{equation}
Here $\frac{\Delta_+ n_i}{n_i}$ is the per-capita rate of territorial acquisition, and $1/d_i$ is the fraction of type $i$ adults surviving to the next iteration. Note that our model of mortality differs from the classic lottery model \citep{chesson_1981}, where mortality affects adults only and occurs after propagule production but before juvenile recruitment. In the latter case, selection on mortality exhibits some density-dependence, although this reflects the fact that newly recruited adults are guaranteed to reproduce before dying, which is not interesting for our purposes here. Our mortality model ensures that selection on $d_i$ is density-independent, allowing us to more clearly separate different sources of density-dependence and density regulation.

@@

Selection in our variable-density lottery model is in general density-dependent, by which we mean that the discrete-time selection factor $(W_2-W_1)/\overline{W}$ from Eq.~\eqref{eq:canonicaldisc} may depend on $N$. More specifically, as we show below, $b$- and $c$- selection are density-dependent, but $d$-selection is not. Note that density-dependent selection is sometimes taken to mean a qualitative change in which types are fitter than others at different densities \citep{travis_2013}. While reversal in the order of fitnesses and co-existence driven by density-regulation are possible in our variable-density lottery (a special case of the competition-colonization trade-off; \citealt{levins_71,tilman_94,bolker_99}), questions related to co-existence are tangential to the aims of the current manuscript and will not be pursued further here.

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{strengthofselection.pdf}
\caption{\label{fig:strengthofselection} Change in the selection coefficient between the beginning and end of a sweep of a type that experiences proportionally $1-\epsilon$ fewer crowding-induced deaths. The population is in demographic equilibrium at the start and end of the sweep.}
\end{figure}

In our variable density lottery, $b$ regulates density and is subject to density-dependent selection, yet $b$-sweeps are qualitatively different from $\delta$ sweeps in the above example. Greater $b$ means not only that more propagules contest the available territories, but also that a greater fraction of unoccupied territories receive propagules. Together, the net density-dependent effect on $b$-selection is negligible: in a single-type equilibrium we have $W_i=1$ and $b_i/\overline{b}=1$, and hence the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ in Eq.~\eqref{eq:masterequalc} has the same value $d_i-1$ at the beginning and end of a $b$-sweep (recall that $\frac{W_2-W_1}{\overline{W}}=\frac{f(\overline{b},N)}{1+f(\overline{b},N)}\frac{b_2-b_1}{\overline{b}}$ for $b$-selection). During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$-sweep (the density-dependent deviation in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for a $b$-sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix C for details). Since selection must already be strong for a $\delta$-sweep to invalidate Eq.~\eqref{eq:canonicalcont}, the density-independent model applies almost exactly for equilibrium $b$-sweeps (Fig.~\ref{fig:bsweep}).

@@

Traits like $b$, $c$ and $d$ will often have pleiotropic interactions which mean that adaptive sweeps will behave similarly to the familiar ``$\delta$ sweeps'' (``Density regulating traits under strong selection''). Thus, our analysis of the variable density lottery does not necessarily imply that Eq.~\eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc} should apply more broadly than previously thought. Rather, $b$, $c$ and $d$ represent a possible set of idealized fitness components mediating the interplay between selection and density in density-regulated populations. The conceptual distinctions they highlight will hopefully be useful regardless of the biological realism of our variable density lottery model.

@@

In the analysis presented here we have restricted our attention to selection in demographically stable populations. The largest deviations from the approximation of density-independent selection (as represented by Eqs.~\eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc}) will likely occur in populations far from demographic equilibrium e.g.~as a result of a temporally-variable environment. This is because extremely strong selection is needed to change population density by an amount comparable to environmental varability (see Fig.~\ref{fig:strengthofselection}). By contrast, temporally-variable environments can dramatically alter frequency trajectories for individual sweeps (e.g. Fig.~9.5 in \cite{otto_2011}; Fig.~5 in \cite{mallet_2012}), as well as the long-term outcomes of selection \citep{lande_2009}.

@@

This pathological behavior occurs because the expected total density of propagules in the respective groups of territories are different: $\langle X \rangle_{p({\mathbf x}|x_1=1,X_1\geq 1)}\approx l_2+1 > \langle X \rangle_{p({\mathbf x}|x_2\geq 2)}\approx l_2 $. As a result, the rare type's behavior is approximated as though it experiences more intense lottery competition than the common type, which cannot be the case since the two types are identical. The effective means must thus be taken in a way that ensures that the expected total propagule density is the same in Eq.~\eqref{eq:meanfieldr} and Eq.~\eqref{eq:meanfielda}.

@@

Having derived the approximation Eq.~\eqref{eq:master}, we now evaluate its domain of validity. Eq.~\eqref{eq:master} relies on ignoring the fluctuations in $x_i$ and $x_j$, such that we can replace them with constant effective mean values. To justify this, we show that the standard deviations $\sigma_{p({\mathbf x}|x_i=1,X_i\geq 1)}(\sum_{j\neq i} c_j x_j)$ and $\sigma_{p({\mathbf x}|x_i\geq 2)}(\sum_j c_j x_j)$ are small compared to the corresponding means $\langle\sum_{j\neq i} c_j x_j\rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ and $\langle\sum_j c_j x_j\rangle_{p({\mathbf x}|x_i\geq 2)}$ in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda}. This result means that using the exact distributions $p({\mathbf x}|x_i=1,X_i\geq 1)$ and $p({\mathbf x}|x_i\geq 2)$ for the effective means would produce an accurate approximation of the components in \eqref{eq:delt_decomp} (though, as we have seen, not a consistent one). It is then clear that the effective means derived in the previous section will also give an accurate approximation since their magnitudes are similar to the exact means; this is obvious from the fact that the expected total number of propagules is of order $\max\{L,2\}$ in both cases.
